Evaluating the Performance and Energy Efficiency of the COSMO-ART model system

### 2014-04-23 trunk/paper_enahpc14/Introduction.tex
adds minor changes; updates paper.bib; includes references for related work

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale  model system  is online  coupled with  the  operational weather
forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
consortium of European weather centers and used in the climate version
(COSMO-CLM) by a wide research community.

The extended  COSMO-ART model  is computationally much  more demanding
than the  COSMO core  since a large  number of additional  tracers and
processes  have  to  be  considered.   Thus this  model  is  currently
severely-limited in  terms of applicability and expensive  in terms of
energy consumption.   Although COSMO has recently been  ported to GPUs
within  the framework of  the High  Performance and  High Productivity
Computing  (HP2C)  Initiative to  optimize  it  for computational  and
energy efficiency (\url{http://www.hp2c.ch/}), significant investments
in  ART  are still  required  to  take it  to  a  similar level.   The
efficiency of COSMO-ART is being addressed in the EU Exa2Green project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

This work is organized as follows: in Sec.~\ref{sec:1} we describe related work, then in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART and specify its  technical setup related to the investigated
performance and energy evaluation methods.  Sec.~\ref{sec:5} describes
the HPC supercomputers used to compute the baseline. Sec.~\ref{sec:3} describes
different  power  measurement  equipment  used  for  conducting  this
benchmarking  study and presents  energy-aware techniques  employed to
determine  the   energy  footprint  of   the  considered  application.
Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
and  highlights areas  where improvements  will be  necessary  for the
subsequent baseline refactoring.  Finally, we conclude with practical
guidelines  to identifying  energy  saving opportunities  and give  an
outlook to future research in Sec.~\ref{concl}.

>>>>

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  these  uncertainties,  the  research  community  is  making  a
concerted international  effort to represent  the underlying physical,
chemical  and aerosol dynamical  processes through  numerical chemical
transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
gases),   developed   at  the   Karlsruhe   Institute  of   Technology
(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
scale model  system is coupled  with the operational  weather forecast
model COSMO \citep{Baldauf-2011}, jointly developed by a consortium of
European weather  centers and used in the  climate version (COSMO-CLM)
by a wide research community.

The extended atmospheric model  COSMO-ART is computationally much more
demanding than  COSMO since a  large number of additional  tracers and
processes  have  to  be  considered.   Thus  the  model  is  currently
severely-limited in  terms of applicability and expensive  in terms of
energy  consumption.   Although  COSMO  has recently  been  ported  to
GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
High Performance and High  Productivity Computing (HP2C) Initiative to
optimise    it    for     computational    and    energy    efficiency
(\url{http://www.hp2c.ch/}), significant investments  in ART are still
required to take  it to a similar level.   The efficiency of COSMO-ART
is     being    addressed    in     the    EU     Exa2Green    project
(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
deliver a  prototype code, which  provides an energy efficiency  of at
least five times of the  baseline value.  Such an implementation would
allow  the  community  to  investigate critical  questions  at  higher
resolution  and   over  longer  periods,   at  reduced  cost   to  the
environment.

This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
introduce  COSMO-ART and specify  its technical  setup related  to the
investigated  performance  and  energy  evaluation  methods.   In  the
subsequent Sec.~\ref{sec:3},  we describe different  power measurement
equipments as well as energy-aware techniques employed to conduct this
benchmarking  study.  Sec.~\ref{sec:4}  presents  the performance  and
energy  requirements  of the  baseline  on  various architectures  and
highlights  areas  where  improvements   will  be  necessary  for  the
subsequent baseline refactoring.   Finally, we conclude with practical
guidelines  to identifying  energy  saving opportunities  and give  an
outlook to future research in Sec.~\ref{concl}.

### 2014-04-23 trunk/paper_enahpc14/Sec4:Evaluation.tex
adds minor changes; updates paper.bib; includes references for related work

In  this  section, we  present  a  performance  and energy  efficiency
evaluation  of  different  achitectures  when  running  the  COSMO-ART
baseline.  We  start by  specifying our measurement  methodology along
with the metrics used to  analyse the results on all platforms.  Then,
we detail the environment  setup gathering the considered HPC systems,
the software  environment and the runtime  configuration.  Finally, we
discuss  benchmark   results  and  power-performance   traces  of  the
application.

\subsection{Measurement methodology}
\label{subsec:4.1}
We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution}       and       \textit{energy-to-solution}.
Time-to-Solution  refers   to  the  total  wall  clock   time  of  the
application execution time. Energy-to-solution is the amount of energy
spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
assessed  by sampling  the instantaneous  peak power  during execution
which  is then  averaged  and multiplied  by  the time-to-solution  to
determine energy-to-solution.   Whenever possible, multiple production
runs of COSMO-ART were  performed to illustrate the reproducibility of
the baseline, and quantify  the significant uncertainties in the power
measurement, as dictated by the available technology.

\subsection{Environment setup}
\label{subsec:4.2}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the milestone
versions   of    COSMO-ART.    In   this    study,   a   complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
v2 ten-core processors operating at 2.2 GHz, themselves connected by a
high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
energy-to-solution benchmark,  a full rack of Monch  constituted of 52
standard  compute   nodes  (monchc[029-080])  was   considered.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
based cluster  used as Piz  Daint pre-post processing cluster.   It is
composed of 42 compute nodes  and has 2 high-speed interconnects based
on FDR:  the first is dedicated to  the MPI traffic and  the second to
the storage high speed traffic.  The 2 login nodes and the 42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
themselves  connected by  a high  speed InfiniBand  network,  based on
Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
was considered.

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both many-core  platforms was  controlled using the  modules framework
which gives  an easy and  flexible mechanism to  access to all  of the
CSCS  provided compilers,  tools  and applications.   For our  initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the  x and  y  coordinates  was defined  by  setting: $nprocx=40$  and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

>>>>

In  this section,  we  describe a  performance  and energy  efficiency
evaluation  of  different  achitectures  when  running  the  COSMO-ART
baseline.  We  start by  specifying our measurement  methodology along
with the metrics used to  analyse the results on all platforms.  Then,
we detail the environment  setup gathering the considered HPC systems,
the  software  environment and  the  run  configuration.  Finally,  we
discuss benchmark  results and  power-performance traces of  the model
system.

\subsection{Measurement methodology}
\label{subsec:4.1}
We  approach  the  assessment  of  the energy  footprint  and  overall
performance    of    COSMO-ART    with    two    important    metrics:
\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
TTS refers to  the total wall clock time  of the application execution
time. ETS  is the amount of  energy spent to  achieve results.  Energy
consumption  is  assessed by  sampling  the  instantaneous peak  power
during execution which  is then averaged and multiplied  by the TTS to
determine  ETS.   Whenever   possible,  multiple  production  runs  of
COSMO-ART  were performed  to  illustrate the  reproducibility of  the
baseline,  and quantify  the  significant uncertainties  in the  power
measurement, as dictated by the available technology.

\subsection{Environment setup}
\label{subsec:4.2}
We have  chosen a state-of-the-art Intel's  third-generation Core (aka
Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
power  measurements,  which  is  slated  to stay  in  service  without
hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
principle at least,  this architecture could be recreated  or found in
an identical configuration beyond  the lifetime of the project.  Since
the baseline benchmark can  be reproduced within an expected variance,
and  that the baseline  run configuration  can be  used in  all future
versions  of the  code, a  fair comparison  will be  made  between the
baseline and  the milestone versions  of COSMO-ART.  In this  study, a
complementary  energy-to-solution benchmarking  comparison  is carried
out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
systems and known to be  more power consuming.  Finally a full tracing
experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
University (UJI) using an  integrated framework, to capture an overall
power profile at a much finer resolution and to get more insights into
the MPI blocking and polling influences on power savings.

\subsubsection{CSCS - Monch}
Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
based  cluster, utilised  by  scientists  who are  part  of the  Swiss
Platform     for      Advanced     Scientific     Computing     (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

\subsubsection{CSCS - Pilatus}
Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
cluster used as Piz Daint pre-post processing cluster.  It is composed
of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
the  first is  dedicated to  the  MPI traffic  and the  second to  the
storage high  speed traffic.   The 2 login  nodes and the  42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
speed, themselves  connected by a high speed  InfiniBand network based
on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
(pilatus[03-44]) was considered.

\subsubsection{UJI - Tintorrum}
Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
For our experiments only a subset of these nodes were considered. This
set is  composed of 16  nodes, each of  which includes two  Intel Xeon
E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
and the platform has 288 GB of DDR3 1333 MHz.

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimisation  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\
